---
title: "Bioregionalization in Ecology using SCCA"
author: "I. Morales-Castilla"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  This vignette walks readers through useful applications of package SCCA in the    field of ecology.
vignette: >
  %\VignetteIndexEntry{usingSCCA}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4, tibble.print_max = 4)
```



This vignette walks readers through useful applications of package scca.R in the field of ecology. Specifically, it does so using a case study for global carnivores, allowing to replicate the results in [van Dam et al](#van_dam).

## Overview

1.  The code installs and load the package.

1.  `sccar.R` applies Correspondance Analysis to an incidence (presence-absence) matrix representing the distribution of the global Carnivora, including both marine and terrestrial species.
    
1.  This vignette walks users through most common functions to obtain bioregions and/or chorotypes and to visualize analytic results. 
    

## Installing the package and R setup

For starters, we will install and load the package. First we do some housekeeping to get R ready.

```{r houskeep, message = FALSE}
## housekeeping
## remove objects and set factors as false
###rm(list=ls())
###options(stringsAsFactors = FALSE)

```

Next, if it is the first time using `scca.R` we need to install the package. Until it is officially published, it can be easily accessed by:

```{r install, message = FALSE, eval = FALSE}

## install and load packages
###library(devtools)
###install_github("UtrechtUniversity/scca", build_vignettes = TRUE)

```

And then it should be loaded.
```{r loadsccar, message = FALSE}
###library(sccar)
###set.seed(0)
```

## Data on global Carnivora
The data has a pre-loaded incidence matrix of presences-absences of global carnivores (Mammalia) from both land and oceans. A quick glance to the data: 

```{r datainspect, message = FALSE, echo = TRUE}
###dim(carnivora)

###head(carnivora[1:6,1:10])

```

The data contains a binary matrix of sites x species (rows x columns) with values of 0 indicating absence and values of 1 indicating presence of a given species _j_ in site _i_. Specifically, there are 41580 sites representing ca. 100 x 100 km grid cells distributed globally and with at least one presence. There are 288 species including marine and terrestrial carnivores.

We can have a look at the geographic distribution of those sites by loading the pre-loaded coordinates of the carnivora dataset, which are named _carnivora_sites_. Also, we can list which species are represented in each column with teh pre-loaded _carnivora_species_.

```{r datainspectmap,, message = FALSE, echo = TRUE,fig.height=5,fig.width=8}
# retrieve coordinates
###sites <- carnivora_sites

# inspect 
###head(sites)

# retrieve species
###carnivores <- carnivora_species

# inspect 
###head(carnivores)

# plot geographic coordinates of sites
###plot(sites$lon,sites$lat,
###     col=adjustcolor('grey',0.4),pch=19,cex=0.3)

```




## Using `scca.R`

Once familiar with the data we can now review basic functions in the package. In this example, we apply Correspondance Analyses aiming to analyze sites according to their species composition (but we could have as well aimed at analyzing species based on their geographic distributions by applying the same procedure to columns instead of rows). The analysis yields a number of clusters in which data can be classified, and which are interpreted as bioregions.

<!--Application of CA reveals that the dataset consists of multiple clusters. As introduced in Section \ref{sec:clusvsord}, we first split the dataset into separate subsets that each consist of a single cluster. These clusters represent sites with similar species composition (termed bioregions). Once we defined the bioregions, we asked whether the sites within them showed any distributional gradient, which would be reflected by the arrangement of sites along “meaningful” CA axes, i.e. axes explaining a considerable amount of variation, as given by the corresponding eigenvalue. If so, it would then be possible to ask whether this gradient would relate to some environmental factor or other spatially patterned process (e.g. distance to main areas of species interchange with other regions \cite{Morales-Castilla2012}). These kinds of questions are common in CA-based ecological investigations and are oftentimes addressed through correlational analyses between meaningful CA-axes and explanatory factors. Since this is well-known practice among ecologists, we did not include this part in the present analysis.-->

scca <- scca_compute(m = carnivora, nstart=100)

## `scca_compute()`: apply CA to a matrix _m_ 

`sccar::scca_compute()` The function scca_compute performs a hierarchical, Spectral Clustering Correspondence Analysis on a matrix representing a bi-partite network. The proces consists of a decomposition of the matrix (svd), a (user-provided) heuristic which transforms the decompostion to input for kmeans clustering (k and the set of observations and, then, the kmeans clustering itself. Each of the resulting clusters (sub-matrices) can be analyzed again untill some stopping condition is met. The output of sccs_compute is a tree (list of lists) in which every node represents one step in the process.

The function to calculate a number of clusters is a heuristic _eigengap_heuristic_, which inspects the eigenspectrum and finds the number of axis before the largest _eigengap_ or distance between each axis and the next.

We can apply the function to the carnivora dataset, generate an object that we call `scca` and inspect the results using the function `scca_print()`.

```{r applyCA, message = FALSE, echo = FALSE}

# application of this function to a large dataset sucha as the carnivora dataset here
# can take up to a minute.
###scca <- scca_compute(m = carnivora, 
###                    iter.max = 10,
###                    nstart = 100,
###                    disconnect.rm = TRUE,
###                    max_eigenvalues = 25,
###                    decomp = "svd",
###                    max_depth = Inf,
###                    heuristic = eigengap_heuristic)

### scca_print(scca , 'k', 'n_labs', 'depth', 'node_type')

```



## `scca_print()`: print hierarchical cluster _m_ 

`sccar::scca_print()` The function scca_prints shows the hierarchical structure of fitted clusters. 



```{r inspectresults, message = FALSE, echo = TRUE}

## inspect the structure of object scca
###scca_print(scca , 'k', 'n_labs', 'depth', 'node_type')

```

There are 17 _leaf nodes_ or clusters identified by the heuristic. The attributes asked to be printed in `scca_print()` have the following meaning: _n_labs_ is the number of obesrvations, _k_ is the number of relevant Eigenvalues (i.e. before the eigengap) at each branch node, _note_type_ is the type of branch (internal) or leaf (terminal) nodes. If no attribute is given, then only the tree hierarchy is printed.





## `scca_get_clusters()`: retrieve and plot or map clusters 

`sccar::scca_get_clusters()` The function `scca_get_clusters` saves the observations to an object, which allows to map obtained clusters or bioregions.



```{r saveplotclusters, message = FALSE, echo = TRUE, fig.height=5,fig.width=8}

## save the clusters in object scca
###clusts<-scca_get_clusters(scca)

## how many clusters?
###length(unique(clusts$cluster))


# plotting clusters
###plot(sites$lon,sites$lat,
###     col=adjustcolor('grey',0.4),pch=19,cex=0.3)

###for(i in sort(unique(clusts$cluster))){#i=2
###points(sites$lon[which(clusts$cluster==i)],
###       sites$lat[which(clusts$cluster==i)],
###       pch=19,cex=0.5,
###       col=adjustcolor(hcl.colors(22)[i],0.6))
###}

```

Regions shaded in different colors represent each cluster or bioregion. The loop above plots all clusters sequentially, but it would be easy to plot any single cluster by adjusting the range of values for _i_.



## `scca_get_node()`: retrieve any given node (axis) and plot its eigen-spectrum

`sccar::scca_get_node()` The function `scca_get_node` retrieves the values for any given node in object scca, which allows to, for example plot eigenspectrums.

There is however, a more direct function to plot the eigen spectrum for any given node: `scca_plot_spectrum()`.


```{r spectralclust, message = FALSE, echo = TRUE}

#node1 <- scca_get_node(scca, node = 1)

## check eigenvalues:

## plot eigenspectrum
###plot(1:length(node1$spectrum),
###     node1$spectrum,ylab="Eigenvalues",xlab="Index")

## the same can be done directly with function scca_plot_spectrum()
###scca_plot_spectrum(scca,1)

```

These plots show the sorted spectrum for all data (node 1), showing two eigenvectors with value equal 1, indicating that there are two fully disconnected clusters. 



## `scca_stability_test()`: run stability test

Tests the stability of an SCCA clustering. Only the clustering at the 'leaves' are considered. After a normal clustering with scca_compute the clustering process is repeated while dropping one of variables from the user provided list. This proces is repeated fro every variable in the list. Every time the clustering output is compared with the original clustering.

The stability measure is the average proportion of overlap (APO: the reverse of APN). APO can be understood as the chance that if 2 observations are in the same cluster in the base clustering they are also in the same cluster in the clustering with 1 column dropped.

```{r stability, message = FALSE, echo = TRUE, fig.height=5,fig.width=8, eval = FALSE}

## check stability (WARNING: takes long time to compute)
###stability <- scca_stability_test(m = carnivora)

# inspect
###stability


```


## `scca_validity_test()`: runs validity test 

Tests the validity of a clustering produced by scca_compute. Computes the three internal validity measures: Connectivity, Silhouette Width and Dunn Index. This function requires a distance matrix, generated by the accessory function `scca_compute_dist()`.

```{r validity, message = FALSE, echo = TRUE, fig.height=5,fig.width=8, eval = FALSE}

##(WARNING: takes long time to compute and uses a lot of memory)

## get distances
###dists <- scca_compute_dist(m = carnivora)


## check stability:
###validity <- scca_valididty_test(scca, dists)

# inspect
###head(validity)


```




### References

<a name="van_dam">
van Dam, Alje, e.a.
</a>
